Binning Strategies and Related Loss for Large Data

### 2015-04-08 Paper.Rnw
New section 5.2 (including a few plots that I don't think will survive much longer than tomorrow's meeting)

\subsection{Frequency Binning Specifications and Frequency Loss}

The loss of frequency information in frequency binning is highly subject to the selection of a discrete color mapping.

@@

freqLossSimDat$FreqBinType <- factor(freqLossSimDat$FreqBinType , levels=c("Standard", "Quantile", "Log Count"))
plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Log Count" & freqLossSimDat$DistType=="Exponential"),]
p1 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = FreqBinType, size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=c(1,3,5,7)) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Freq Loss") + 
  theme(legend.position="bottom")
plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Log Count" & freqLossSimDat$DistType=="Normal"),]
p2 <-  qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = FreqBinType, size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=c(1,3,5,7)) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Freq Loss") + 
  theme(legend.position="bottom")
plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Log Count" & freqLossSimDat$DistType=="Uniform"),]
p3 <-qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = FreqBinType, size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=c(1,3,5,7)) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Freq Loss") + 
  theme(legend.position="bottom")

>>>>

\subsection{Frequency Binning Specifications and Frequency Loss}
\ktm{
The reduced binned data from spatial binning contains the center and count information for all bins.  The bin frequencies may themselves be binned in order to discretize the color scale for the binned scatterplot. The loss of frequency information in frequency binning is dependent on the selection of a discrete color mapping, where the binning algorithm and number of frequency bins must be specified. To explore frequency loss characteristics using the simulated bivariate data sets, each of the three fine resolution sets were first spatially binned using standard rectangular binning to produce three sets of reduced binned data; then standard, quantile and log frequency binning algorithms were applied with various numbers of frequency bins. As discussed in Section~\ref{FreqLoss}, we will limit the considered options to using between one and seven frequency bins due to the upper limits of human perceptual abilities \citep{colorperception}.

Figure~\ref{fig:FreqLossCompare} displays the frequency loss from using standard and quantile frequency binning algorithms, with between one and seven frequency bins, from each set of simulated data. We first notice that large difference in the magnitude of frequency losses based on the bivariate distribution; frequency losses are highest for the exponential data, lower for the normal data and lowest for the uniform data. We also note that the frequency loss is a decreasing function of the number of frequency bins for both binning algorithms, thus we should consider using the maximum number of bin shades that human perception is able to simultaneously discern.  Lastly, the frequency loss with the standard frequency binning may be higher than for the quantile-based algorithm when a minimal number of bins are used; but in all three simulated data sets, this is reversed when seven frequency bins are used. Thus we recommend using the standard algorithm over the quantile-based algorithm for frequency binning because it had smaller frequency loss at the recommended seven bins, and additionally carries the benefit that equally dividing the bin count range into equally spaced bins creates a simpler shading interpretion than dividing by quantiles.

Due to the difference in scales between counts and log counts, the frequency loss and log frequency loss can not be directly compared. Frequency binning using the log counts did however behave very similarly, where log frequency loss decreased as the number of frequency bins increased.  Thus the same advise to use seven frequency bins holds for shading based on log counts as well. 
}

<<FreqLossDisplay,echo=F,include=F,eval=T,fig.width=15, fig.height=5, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="I am not sure that this figure is needed to demonstrate frequency binning. If so, do the colors between freq bins need to be made more distinguishable? i.e. in the quantile binning the bins starting at 1, 8 and 36.5 are hard to differentiate because the shade is determined by the breaks. ">>= 
binout1 <- RectBin2d(datlist[[2]]$x,datlist[[2]]$y,0,0,5,5,type="standard")
binoutFreqGroups <- freqBin(binout1, binType="standard", ncolor=5, logCount=FALSE)
binoutQuantFreqGroups <- freqBin(binout1, binType="quantile", ncolor=5, logCount=FALSE)
binoutLogFreqGroups <- freqBin(binout1, binType="standard", ncolor=5, logCount=TRUE)

# Plot with Orginal Frequencies
p1 <- qplot(binxs, binys, geom="tile", fill=binfreq, data=binoutFreqGroups[[1]],
            main="No Freq Binning") +
  xlab("") + ylab("")+
  scale_fill_gradient(low="#56B1F7", high="#132B43") + 
  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
                     panel.grid= element_blank())
# Plot with Binned Frequencies
p2 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutFreqGroups[[1]],
            main="Standard Freq Binning") +
    xlab("") + ylab("")+
  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
                      breaks=unique(binoutFreqGroups[[1]]$freqgroup)) + 
  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
                     panel.grid= element_blank())
# Plot with Quantile Binned Frequencies
p3 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutQuantFreqGroups[[1]],
            main="Quantile Freq Binning") +
    xlab("") + ylab("")+
  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", breaks=unique(binoutQuantFreqGroups[[1]]$freqgroup)) + 
  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
                     panel.grid= element_blank())
# Plot with LogCount Binned Frequencies
p4 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutLogFreqGroups[[1]],
            main="Log Freq Binning") +
    xlab("") + ylab("")+
  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", breaks=unique(binoutLogFreqGroups[[1]]$freqgroup)) + 
  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
                     panel.grid= element_blank())

grid.arrange(p1,p2,p3,p4, nrow=1)
@

@@

freqLossSimDat$FreqBinType <- factor(freqLossSimDat$FreqBinType , levels=c("Standard", "Quantile", "Log Count"))
plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Log Count" & freqLossSimDat$DistType=="Exponential"),]
p1 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = FreqBinType, size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Freq Loss") + 
  theme(legend.position="bottom")
plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Log Count" & freqLossSimDat$DistType=="Normal"),]
p2 <-  qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = FreqBinType, size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Freq Loss") + 
  theme(legend.position="bottom")
plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Log Count" & freqLossSimDat$DistType=="Uniform"),]
p3 <-qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = FreqBinType, size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Freq Loss") + 
  theme(legend.position="bottom")

<<FreqLossCompare2,echo=F,include=T,eval=T,fig.width=10, fig.height=4, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total log frequency loss from log frequency binning of reduced binned data from each bivariate distribution. I don't think this plot will make the final cut, but it is interesting that after logging the bin counts, the log freq loss for the uniform data does not improve with more bins (might as well use one shade for all bins)">>= 
### Loop over ncolors 1 to 7 and also over data distribution
# find freq loss from standard, quantile, logCount
freqLossSimDat <- NULL
distrib <- c("Uniform", "Normal", "Exponential")
for (distidx in 1:3){
  binout1 <- RectBin2d(datlist[[distidx]]$x,datlist[[distidx]]$y,0,0,5,5,type="standard")
  for (ncol in 1:7){
    binoutFreq <- rbind(freqBin(binout1, binType="standard", ncolor=ncol, logCount=FALSE)[[2]],
                        freqBin(binout1, binType="quantile", ncolor=ncol, logCount=FALSE)[[2]], 
                        freqBin(binout1, binType="standard", ncolor=ncol, logCount=TRUE)[[2]])
    binoutFreq$FreqBinType <- c("Standard", "Quantile", "Log Count")
    binoutFreq$DistType <- distrib[distidx]
    binoutFreq$ncolor <- ncol 
    freqLossSimDat <- rbind(freqLossSimDat,binoutFreq)
  }
}

freqLossSimDat$FreqBinType <- factor(freqLossSimDat$FreqBinType , levels=c("Standard", "Quantile", "Log Count"))
plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType=="Log Count" & freqLossSimDat$DistType=="Exponential"),]
p1 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = I(3), size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
  theme(legend.position="bottom")
plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType=="Log Count" & freqLossSimDat$DistType=="Normal"),]
p2 <-  qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = I(3), size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
  theme(legend.position="bottom")
plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType=="Log Count" & freqLossSimDat$DistType=="Uniform"),]
p3 <-qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = I(3), size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
  theme(legend.position="bottom")
